Forecasting Gamma-Ray Bursts using Gravitational Waves

### 2018-11-18 Paper/Manuscript.tex
Changes for the referee report

In a network of $N$ interferometers, the total SNR is defined as follows \cite{Finn:2000hj}
%%
%
\be
\rho_\text{tot} = \left[{\sum_{i=1}^N \rho^2_i}\right]^{1/2}\, , \label{eq:SNR_total}
\ee
%
%%
where Eq.~(\ref{eq:SNR_v2}) is employed to compute $\rho^2_i$ of individual interferometers each with a specific quality factor $Q_i$.
Here, instead of working with specific quality factors of the LIGO and the Virgo interferometers, we consider
a \emph{model network} of three IFOs ($N=3$), where one is near-optimally oriented/positioned with $|Q|=2\sigma=4/5$.
\todo{ADD some discussion here as to why we don't use specific LV Q's}
The second IFO operates at the RMS average; thus it has $|Q|=2/5$. %compared with the optimal $|Q| =1$ case.
It remains to assign a quality factor to the third IFO. We wish to make this operate below the RMS average, % in terms of its orientation and the sky location of the sources,
but it would be pointless to simply set $Q=0$. We therefore pick the values $\iota=0$ (i.e, $|h_+|=|h_\times|$)
and $\psi=0$ for simplicity %(aligning the $x$-axis of the ``$+$'' with the of the IFO
then introduce the following $\{\theta,\phi\}$-only averages
%
\be
 \langle F_+^2\rangle_{\theta,\phi} = \frac{7}{30},\qquad \langle F_\times^2\rangle_{\theta,\phi} = \frac{1}{6}\, 
\ee
yielding
%
\be
\langle |Q(\theta,\phi,\psi=0,\iota=0)|^2 \rangle_{\theta,\phi} = \f{2}{5} \label{eq:RMS_th_phi}.
\ee
Using this value, we assign our below-average third IFO %$\langle|Q|^2\rangle^{1/2}
$|Q|=4/5-\sqrt{2/5}\simeq 0.1675$.
%which is essentially a $\{\theta$-$\phi\}$-RMS awayfrom the overall RMS averaged value of 2/5.
This may seem somewhat ad hoc, 
but it is our way of representing the sub-RMS-average response of the third IFO due to its sub-optimal configuration. 
One may very well use another scheme to generate this below-average response.

@@

\begin{table}[h]
\centering
\begin{tabular}{ccccc}
\hline\hline
 IFO$_i$  %& $\langle |Q_i|^2\rangle$ 
 & $|Q_i|$ & $\tilde{A}_i$ & $\tilde{H}_i(f)$ & IFO response\T\B \\
 %\hline
 \hline
 1 &  $\tfrac{4}{5} $&  $\tfrac{4}{5}A$ & $\tfrac{4}{5}A h_0\, f^{-7/6}$ & high ($h$) \T\\
 2 &  $\tfrac{2}{5} $ & $\tfrac{2}{5}A$ & $ \tfrac{2}{5}A h_0 \, f^{-7/6}$ & medium ($m$)\B\T \\
 3 &  $\quad \tfrac{4}{5}-\sqrt{\tfrac{2}{5}} \quad $
 & $\quad\left(\tfrac{4}{5}-\sqrt{\tfrac{2}{5}}\right)A \quad$ & $\quad \left(\tfrac{4}{5}-\sqrt{\tfrac{2}{5}}\right)A h_0 \, f^{-7/6}\quad$& low ($l$) \B \\
 \hline\hline
\end{tabular}
\caption{%The averaged quality factors $\langle|Q|^2\rangle$ of 
Our model for the Advanced-LIGO-Virgo network ca. 2020. %of three L-shaped interferometers (IFOs). 
Column two shows the quality factor $Q_i$ for each IFO and column three the corresponding amplitude scalar $\tilde{A}_i$ introduced
in Eq.~(\ref{eq:Hi_s}).
Column four lists the magnitude of the frequency-domain strain at each IFO via Eq.~(\ref{eq:Hi_s}).
The last column states each IFO's response, e.g., high means high strain, thus high SNR.
%shows how loud the GW strain is at a given IFO due source sky position and IFO orientation. 
%oriented/positioned an IFO is with respect to a compact binary source. 
%High is $2\sigma$ above the average as explained above.
%Medium is the RMS-averaged (one $\sigma$) response. 
Recall $A= \pi^{-2/3} \sqrt{5/24}$. Maximum value for $|Q_i|$ is 1 and minimum is 0.
%Note that for an interferometer where the angle between the arms is $\alpha$, $A$ rescales to $A\sin\alpha$.
}\label{table:network_params}
\end{table}
%
%
We can now write the total SNR for our three L-IFO network
%%
%
\begin{align}
\rho_\text{tot} &= \sqrt{\rho^2_h+\rho^2_m+\rho^2_l} \label{eq:SNR_total_3IFOs} 
%	      &= 2A h_0 \sqrt{1+\tfrac{2}{5}+\tfrac{4}{25}} \left[\int_0^{f_\text{ISCO}} df\, \f{f^{-7/3}}{S_\text{noise}(f)}\right]^{1/2}\nn\\
%	      &= \sqrt{\frac{13}{10}}\,\pi^{-2/3}\,\f{c}{D}\left(\f{G M_c}{c^3}\right)^{5/6}\left[\int_0^{f_\text{ISCO}} df\, \f{f^{-7/3}}{S_\text{noise}(f)}\right]^{1/2} \label{eq:SNR_total_3IFOs_v2}.
\end{align}
%
%%
with $\rho_h, \rho_m,\rho_l$ given by Eq.~(\ref{eq:SNR_i}) and Table~\ref{table:network_params}.
%Although we made several simplifying assumptions to obtain $\rho_\text{tot}$, these should not affect our total SNR estimation too much.
%For example, averaging over all the angles underestimates the SNR of an interferometer with a near-ideal orientation with respect to the incoming GWs
%while also overestimating the SNR of a badly oriented detector. Rewriting Eq.~(\ref{eq:SNR_total}) as Eq.~(\ref{eq:SNR_total_3IFOs}) compensates
%for these.
%
%In Sec.~\ref{sec:GW170817} we test our model network with GW170817 whose signal got picked up by the LIGO detectors around $f=30$\,Hz.
%To do so we will change the lower limit of the integral in Eq.~(\ref{eq:SNR_total_3IFOs}) from $0$ to $30$.
%For our advance warning times we will further use separate initial frequencies for each inteferometer as we will explain in Sec.~\ref{sec:results}.

@@

the fastest observed NS spin in a BNS system is $\lesssim 0.04$ \cite{0264-9381-26-7-073001}.

@@

We use a $\verb|Mathematica|$ based code developed at UCD to do the same for GW170817.

@@

\bar\rho_h = 29.2,\quad \bar\rho_m = 14.6,\quad \bar\rho_l = 6.11 \label{eq:GW170817_SNRs}.

@@

\bar\rho_\text{tot}= \sqrt{\bar\rho^2_h+\bar\rho^2_m+\bar\rho^2_l}= 32.6 \, \label{eq:SNR_total_GW170817} .

>>>>

In a network of $N$ interferometers, the total SNR is defined as follows \cite{Finn:2000hj}
%%
%
\be
\rho_\text{tot} = \left[{\sum_{i=1}^N \rho^2_i}\right]^{1/2}\, , \label{eq:SNR_total}
\ee
%
%%
where Eq.~(\ref{eq:SNR_v2}) is employed to compute $\rho^2_i$ of individual interferometers each with a specific quality factor $Q_i$.
Here, instead of working with specific quality factors of the LIGO and the Virgo interferometers, we consider
a \emph{model network} of three IFOs ($N=3$), where one operates at the RMS average; thus it has $|Q|=2/5$. 
The second IFO is near-optimally oriented/positioned so we assign it a quality
factor one $\sigma$ above the average: $|Q|=2/5+\sigma=4/5$.
It remains to assign a quality factor to the third IFO. We wish to make this operate below the RMS average, % in terms of its orientation and the sky location of the sources,
but it would be pointless to simply set $Q=0$. We therefore pick the values $\iota=0$ (i.e, $|h_+|=|h_\times|$)
and $\psi=0$ for simplicity %(aligning the $x$-axis of the ``$+$'' with the of the IFO
then introduce the following $\{\theta,\phi\}$-only averages
%
\be
 \langle F_+^2\rangle_{\theta,\phi} = \frac{7}{30},\qquad \langle F_\times^2\rangle_{\theta,\phi} = \frac{1}{6}\, 
\ee
yielding
%
\be
\langle |Q(\theta,\phi,\psi=0,\iota=0)|^2 \rangle_{\theta,\phi} = \f{2}{5} \label{eq:RMS_th_phi}.
\ee
Using this value, we assign our below-average third IFO %$\langle|Q|^2\rangle^{1/2}
$|Q|=4/5-\sqrt{2/5}\simeq 0.1675$.
%which is essentially a $\{\theta$-$\phi\}$-RMS awayfrom the overall RMS averaged value of 2/5.
This may seem somewhat ad hoc, 
but it is our way of representing the sub-RMS-average response of the third IFO due to its sub-optimal configuration. 
One may very well use another scheme to generate this below-average response.
Given that $\rho_\text{tot}$ is a sum of quadratures the lowest SNR contributes a small percentage to this [cf. Eq.~(\ref{eq:SNR_total})].

@@

\begin{table}[h]
\centering
\begin{tabular}{ccccc}
\hline\hline
 IFO$_i$  %& $\langle |Q_i|^2\rangle$ 
 & $|Q_i|$ & $\tilde{A}_i$ & $\tilde{H}_i(f)$ & IFO response\T\B \\
 %\hline
 \hline
 1 &  $\tfrac{4}{5} $&  $\tfrac{4}{5}A$ & $\tfrac{4}{5}A h_0\, f^{-7/6}$ & high ($h$) \T\\
 2 &  $\tfrac{2}{5} $ & $\tfrac{2}{5}A$ & $ \tfrac{2}{5}A h_0 \, f^{-7/6}$ & medium ($m$)\B\T \\
 3 &  $\quad \tfrac{4}{5}-\sqrt{\tfrac{2}{5}} \quad $
 & $\quad\left(\tfrac{4}{5}-\sqrt{\tfrac{2}{5}}\right)A \quad$ & $\quad \left(\tfrac{4}{5}-\sqrt{\tfrac{2}{5}}\right)A h_0 \, f^{-7/6}\quad$& low ($l$) \B \\
 \hline\hline
\end{tabular}
\caption{%The averaged quality factors $\langle|Q|^2\rangle$ of 
Our model for the Advanced-LIGO-Virgo network ca. 2020. %of three L-shaped interferometers (IFOs). 
Column two shows the quality factor $Q_i$ for each IFO and column three the corresponding amplitude scalar $\tilde{A}_i$ introduced
in Eq.~(\ref{eq:Hi_s}).
Column four lists the magnitude of the frequency-domain strain at each IFO via Eq.~(\ref{eq:Hi_s}).
The last column states each IFO's response, e.g., high means high sensitivity, thus high SNR.
%shows how loud the GW strain is at a given IFO due source sky position and IFO orientation. 
%oriented/positioned an IFO is with respect to a compact binary source. 
%High is $2\sigma$ above the average as explained above.
%Medium is the RMS-averaged (one $\sigma$) response. 
Recall $A= \pi^{-2/3} \sqrt{5/24}$. Maximum value for $|Q_i|$ is 1 and minimum is 0.
%Note that for an interferometer where the angle between the arms is $\alpha$, $A$ rescales to $A\sin\alpha$.
}\label{table:network_params}
\end{table}
%
%
We can now write the total SNR for our three L-IFO network
%%
%
\begin{align}
\rho_\text{tot} &= \sqrt{\rho^2_h+\rho^2_m+\rho^2_l} \label{eq:SNR_total_3IFOs} 
%	      &= 2A h_0 \sqrt{1+\tfrac{2}{5}+\tfrac{4}{25}} \left[\int_0^{f_\text{ISCO}} df\, \f{f^{-7/3}}{S_\text{noise}(f)}\right]^{1/2}\nn\\
%	      &= \sqrt{\frac{13}{10}}\,\pi^{-2/3}\,\f{c}{D}\left(\f{G M_c}{c^3}\right)^{5/6}\left[\int_0^{f_\text{ISCO}} df\, \f{f^{-7/3}}{S_\text{noise}(f)}\right]^{1/2} \label{eq:SNR_total_3IFOs_v2}.
\end{align}
%
%%
with $\rho_h, \rho_m,\rho_l$ given by Eq.~(\ref{eq:SNR_i}) and Table~\ref{table:network_params}.
%Although we made several simplifying assumptions to obtain $\rho_\text{tot}$, these should not affect our total SNR estimation too much.
%For example, averaging over all the angles underestimates the SNR of an interferometer with a near-ideal orientation with respect to the incoming GWs
%while also overestimating the SNR of a badly oriented detector. Rewriting Eq.~(\ref{eq:SNR_total}) as Eq.~(\ref{eq:SNR_total_3IFOs}) compensates
%for these.
%
%In Sec.~\ref{sec:GW170817} we test our model network with GW170817 whose signal got picked up by the LIGO detectors around $f=30$\,Hz.
%To do so we will change the lower limit of the integral in Eq.~(\ref{eq:SNR_total_3IFOs}) from $0$ to $30$.
%For our advance warning times we will further use separate initial frequencies for each inteferometer as we will explain in Sec.~\ref{sec:results}.

Let us conclude this subsection with a brief discussion on 
%our particular choice to represent the sensitivities of the near-future ALV network.
an alternate approach to construct a model for the near-future ALV network.
This involves selecting one IFO as the ``base'' detector, e.g., LIGO Livingstone then running Monte-Carlo simulations over the four random angles $\{\theta,\phi,\psi,\iota\}$. 
We would subsequently compute the quality factor of the base IFO using the
simulation results for the four angles.
We can then obtain the quality factors of the remaining two IFOs 
--- in this particular case, LIGO Hanford and Virgo --- via their positions and orientations with respect to the base IFO. Given that the four angles
are completely random, there should be no preferred point in the angle parameter space, thus the result of the Monte-Carlo runs should converge
to the RMS average. Correspondingly, we would expect one of the other two IFOs to have somewhat higher sensitivity, and the other one to have lower sensitivity.
This network can be further expanded using exact positions of LIGO India
and KAGRA detectors. The expansion of the detector network from three to five IFOs will offer significant improvements in source localization.
However, given that the total network SNR is a sum of quadratures, the 
expanded network will not accumulate significantly higher SNR than
our model ALV network. As such our model network should serve as a good proxy for the advance warning time computations.

@@

the fastest observed NS spin in a BNS system is $\lesssim 0.04-0.05$ \cite{0264-9381-26-7-073001, Stovall:2018ouw}.

@@

We use a $\verb|Mathematica|$ based code developed at University College Dublin (UCD) to do the same for GW170817.

@@

\bar\rho_h = 28.7,\quad \bar\rho_m = 14.3,\quad \bar\rho_l = 6.00 \label{eq:GW170817_SNRs}.

@@

\bar\rho_\text{tot}= \sqrt{\bar\rho^2_h+\bar\rho^2_m+\bar\rho^2_l}= 32.0 \, \label{eq:SNR_total_GW170817} .
